F\# Data: \textnormal{Making structured data first-class citizens}

### 2015-11-17 pldi-2016/paper.tex
More edits and add practical experience

\documentclass[10pt,preprint,clearpagebib]{sigplanconf}

\usepackage{listings}
\usepackage{enumitem}
\usepackage[hyphens]{url}
\usepackage[svgnames]{xcolor}
\definecolor{lcl}{RGB}{160,0,128}
\usepackage[colorlinks=true,breaklinks,draft=false]{hyperref}
\hypersetup{urlcolor=lcl,linkcolor=lcl}
\newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}

\usepackage{amssymb}
\usepackage{multicol}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage[fleqn]{amsmath}
\usepackage{graphics} 
\usepackage{stmaryrd}
\usepackage{amsthm}
\usepackage{ftnright}
\usepackage[T1]{fontenc}
\usepackage{semantic}
\usepackage{enumitem}
\usepackage[hang,flushmargin]{footmisc} 
\usepackage[notquote]{hanging}
\usepackage{flushend}

@@

% TODO: Adapt XML example 
% TODO: Switch from sum types to variant types
% TODO: Discuss 'stability' of the inference (minor change in schema does not cause dramatic cahnges)
% TODO: Say that any variant is a subtype of any other variant 
% TODO: Say that anything is a subtype of variant
% TODO: Say that variant is a last-resort thing
% TODO: Double check that bottom/top are used correctly
% TODO: Explain (even more) that there is no least upper bound, which is why Theorem 3 is good enough
% TODO:  -- but maybe we can have lub w.r.t variants? or something?
% TODO: Avoid "no other rule applies" in the formal system
% TODO: Practical evaluation (lots of downloads, lots of Github - how often people customize samples?)
% TODO: Explain why we need option type (because T + null is a variant)

% TODO: In the PLDI version, we delete all the unnecessary runtime conversions
% but we should say that those happen for practical reasons (e.g. float -> int, or anyhing to string)

% T :> S means that S is a subtype of T (and S can be used where T is expected)
% TOP is the universal supertype
% BOTTOM is the universal subtype

% variant is the top type 
%  (variant is a suprtype of everything; everything is a subtype of variant; 
%   everything can be used where variant is expected)
%
% bottom is the bottom type
%  (bottom is a subtype of everything; everything is a supertype of bottom
%   bottom can be used where anything else is expected - but it has no values)
%
% define :=: when <: and :>, then take equivalence classes

% least upper bound of T1 and T2 is a supertype of both (T :> T1 and T :> T2)
% and for all S such that S :> T1 and S :> T2, it holds that S :> T

@@

\begin{abstract}
Most modern applications interact with external services and access data in formats such as 
XML, JSON and CSV. Static type systems do not understand external data sources and only 
make accessing data more cumbersome. Should we give up and leave the messy world of external data 
to dynamic typing and runtime checks? Of course, not!

@@

For example, OpenWeatherMap provides an end-point for getting current weather for a 
given city\footnote{See ``Current weather data'': \url{http://openweathermap.org/current}}. 
The documentation describes the URL parameters and shows one sample JSON to illustrate the typical
response structure. Using a standard library for working with JSON and HTTP, we might call the 
service and read the temperature as follows:
%
\begin{equation*}
\begin{array}{l}
 \kvd{let}~\ident{doc}=\ident{Http.Request}(\str{http://weather.org/?q=NYC}) \\
 \kvd{match}~\ident{JsonValue.Parse}(\ident{doc})~\kvd{with} \\
 |~\ident{Record}(\ident{root})\rightarrow \\
 \quad \kvd{match}~\ident{Map.find}~\str{main}~\ident{root}~\kvd{with} \\
 \quad |~\ident{Record}(\ident{main})\rightarrow \\
 \quad \quad \kvd{match}~\ident{Map.find}~\str{temp}~\ident{main}~\kvd{with} \\
 \quad \quad |~\ident{Number}(\ident{num})\rightarrow \ident{printfn}~\str{Nice \%f degrees!}~\ident{num} \\
 \quad \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\
 \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\
 |~\_\rightarrow \ident{failwith}~\str{Incorrect format} 
\end{array}
\end{equation*}
%
The code assumes that the response has a particular format described in the documentation. The
root node must be a record with a \str{main} field, which has to be another record containing
a numerical \str{temp} field. When the format is incorrect, the data access simply fails
with an exception.

Using the JSON type provider from F\# Data, we can write code with exactly the 
same functionality in two lines:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{W} = \ident{JsonProvider}\langl\str{http://weather.org/?q=NYC}\rangl \\[0.1em]
 \ident{printfn}~\str{Nice \%f degrees!}~(\ident{W.GetSample().Main.Temp})
\end{array}
\end{equation*}
%
Here, $\ident{JsonProvider}\langl\str{...}\rangl$ invokes a type provider at 
compile-time with the URL as a sample. The type provider infers the structure of the response
and provides a type with a \ident{GetSample} method that returns a parsed JSON with nested
properties \ident{Main.Temp}, returning the temperature as a number.

@@

% Although the F\# Data library \cite{fsharp-data} has been widely adopted and is one of the most downloaded F\# libraries 
% it has not yet been presented in an academic form. Additional documentation and source code can 
% be found at \url{http://fsharp.github.io/FSharp.Data}.

@@

We start with an informal overview that shows how F\# Data type providers simplify working with 
JSON, XML and CSV. We also introduce the necessary aspects of F\# 3.0 type providers along the 
way. The examples in this section illustrate a number of key properties of our type inference 
algorithm:

\begin{itemize}
\item Our mechanism is robust and predictable. This is important as the user directly works with 
  the inferred types and should understand why a specific type was inferred from a given 
  sample\footnote{In particular, we do not use probabilistic methods where adding additional 
  sample could change the shape of the type.}.

\item Our inference mechanism prefers records over unions. This better supports developer tooling --
  most F\# editors provide code completion hints on ``.'' and so types with properties (records)
  are easier to use than types that require pattern matching.

\item Finally, we handle a number of practical concerns that may appear overly complicated, but
  are important in the real world. This includes support for different numerical types, \kvd{null}
  values and missing data and also different ways of representing Booleans in CSV files.
\end{itemize}

@@

The JSON format used in the example in Section~\ref{sec:introduction} is a popular format for data
exchange on the web based on data structures used in JavaScript. The following is the definition 
of the \ident{JsonValue} type used earlier to represent parsed JSON:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{JsonValue} = \\[0.1em]
 \quad|~\ident{Null} \\[0.1em]
 \quad|~\ident{Number}~\kvd{of}~\ident{decimal} \\[0.1em]
 \quad|~\ident{String}~\kvd{of}~\ident{string} \\[0.1em]
 \quad|~\ident{Boolean}~\kvd{of}~\ident{bool} \\[0.1em]
 \quad|~\ident{Record}~\kvd{of}~\ident{Map}\langl\ident{string}, \ident{JsonValue}\rangl \\[0.1em]
 \quad|~\ident{Array}~\kvd{of}~\ident{JsonValue}[] \\[0.1em]
\end{array}
\end{equation*}
%
The OpenWeatherMap example in the introduction used only a (nested) record containing a numerical
value. To demonstrate other aspects of the JSON type provider, we look at a more complex example
that also involves \kvd{null} value and an array:
%
{\small{
\begin{verbatim}
  [ { "name": "Jan", "age": 25 },
    { "name": "Alexander", "age": 3.5 },
    { "name": "Tomas" } ]
\end{verbatim}
}}
%
\noindent
Say we want to print the names of people in the list with an age if it is available. As before, 
the standard approach would be to pattern match on the parsed \ident{JsonValue}. The code would 
check that the top-level node is a \ident{Array}, iterate over the elements checking that each is
a \ident{Record} with certain properties and throw an exception or skip values in incorrect format.
The standard approach can be made nicer by defining helper functions. However, we still need to
specify names of fields as strings, which is error prone and can not be statically checked.

Assuming the \strf{people.json} file contains the above example and \ident{data} is a string value
that contains another data set in the same format, we can print names and ages as follows:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{People}~=~\ident{JsonProvider}\langl\str{people.json}\rangl\hspace{1em} \\[0.6em]
 \kvd{let}~\ident{items}~=~\ident{People.Parse}(\ident{data})\\[0.1em]
 \kvd{for}~\ident{item}~\kvd{in}~\ident{items}~\kvd{do}\\[0.1em]
 \quad\ident{printf}~\str{\%s }~\ident{item.Name}\\[0.1em]
 \quad\ident{Option.iter}~(\ident{printf}~\str{(\%f)})~\ident{item.Age}
\end{array}
\end{equation*}
%
The code achieves the same simplicity as when using dynamically typed languages, but it is statically 
type-checked. In contrast to the earlier example, we now use a local file \strf{people.json} as a 
sample for the type inference, but then processes data from another source.

\paragraph{Type providers.}
The notation $\ident{JsonProvider}\langl\str{people.json}\rangl$ on the first line passes a 
\emph{static parameter} to the type provider. Static parameters are resolved at compile-time, 
so the file name has to be a constant. The provider analyzes the sample and generates
a type that we name \ident{People}. In editors that use the F\# Compiler Service, the 
type provider is also executed at development-time and so the same provided types are
used in code completion.

The \ident{JsonProvider} uses a type inference algorithm (Section~\ref{sec:inference})  and
infers the following types from the sample:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{Entity}~=  \\[0.1em]
 \quad \kvd{member}~\ident{Name}~:~\ident{string} \\[0.1em]
 \quad \kvd{member}~\ident{Age}~:~\ident{option}\langl \ident{decimal}\rangl \\[0.6em]
 \kvd{type}~\ident{People}~=  \\[0.1em]
 \quad \kvd{member}~\ident{GetSample}~:~\ident{unit}~\rightarrow~\ident{Entity}[] \\[0.1em]
 \quad \kvd{member}~\ident{Parse}~:~\ident{string}~\rightarrow~\ident{Entity}[] \\[0.1em]
\end{array}
\end{equation*}
%
The type \ident{Entity} represents the person. The field \ident{Name} is available for all
sample values and is inferred as \ident{string}. The field \ident{Age} is marked as optional,
because the value is missing in one sample. The two sample ages are an integer $25$ and a 
decimal $3.5$ and so the common inferred type is \ident{decimal}.

The type \ident{People} provides two methods for reading data. \ident{GetSample} returns the
sample used for the inference and \ident{Parse} parses a JSON string containing data in the same 
format as the sample. Since the sample JSON is a collection of records, both methods return an 
array of \ident{Entity} values.

\paragraph{Error handling.}
In addition to the structure of the types, the type provider also specifies what code should be 
executed at run-time in place of \ident{item.Name} and other operations. This is done through a 
\emph{type erasure} mechanism demonstrated in Section~\ref{sec:providers-csv}. In this example,
the runtime behaviour is the same as in the hand-written sample in Section~\ref{sec:introduction} --
a member access throws an exception if the document does not have the expected format.

Informally, the safety property discussed in Section~\ref{sec:safety} states that if the inputs
are subtypes of some of the provided samples (\emph{i.e.}~the samples are representative), then
no exceptions will occur. In other words, we cannot avoid all failures, but we can prevent some.
If the OpenWeatherMap changes the response format, the sample in Section~\ref{sec:introduction}
will not re-compile and the user knows that the code needs to be changed. What this means for
the traditional ML type safety is discussed in Section~\ref{sec:safety-discuss}.

\paragraph{The role of records.}
The sample code is easy to write thanks to the fact that most F\# editors provide code completion
when ``.'' is typed. The developer does not need to look at the sample JSON file to see what fields
are available for a person. To support this scenario, our inference algorithm prefers records
(we also treat XML elements and CSV rows as records).

@@

\subsection{Processing XML data, alternative}
\label{sec:providers-xml}

In XML, documents are formed by (possibly) nested elements with attributes. We can view elements
as records (unlike in JSON, the records are named) with a field for each attribute and an additional
special field for the nested contents (which may be a collection of elements).

As an example, consider a simple extensible document format where a root element
{\ttfamily\small <doc>} can contain a number of document element, one of which is
{\ttfamily\small <heading>} representing headings:
%
{\small{
\begin{verbatim}
  <doc>
    <heading>Working with JSON</heading>
    <p>Type providers make this easy.</p>
    <heading>Working with XML</heading>
    <p>Processing XML is as easy as JSON.</p>
    <image source="xml.png" />
  </doc>
\end{verbatim}
}}
%
\noindent
The F\# Data library has been primarilly designed to allow easy reading of data. For example,
say we want to print all headings that appear in the document. The sample shows a part of the 
possible document structure (in particular the {\ttfamily\small <heading>} element we want to
read), but it does not show all possible elements (say, {\ttfamily\small <table>}). Assuming the
above document is \strf{sample.doc}, we can write:
%
\noindent
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{Document}~=~\ident{XmlProvider}\langl\str{sample.doc}\rangl\hspace{1em} \\[0.5em]
 \kvd{let}~\ident{root}~=~\ident{Document.Load}(\str{http://.../another.doc})\\
 \kvd{for}~\ident{elem}~\kvd{in}~\ident{root.Doc}~\kvd{do}\\
 \quad\kvd{match}~\ident{elem.Heading}~\kvd{with} \\
 \quad|~\ident{Some text} \rightarrow \ident{printf}~\str{ - \%s }~\ident{text}\\
 \quad|~\ident{None} \rightarrow ()\\
\end{array}
\end{equation*}
%
The example iterates over a collection of elements \ident{elem} returned by \ident{root.Doc}. 
Our system provides a typed access to elements known from the sample document and so we can
write \ident{elem.Heading}.

\paragraph{Open world and variant types.}
In the above example, The system does not infer a union type representing a choice between
heading, paragraph and image. By its nature, the XML format is extensible and the sample cannot 
include all possible examples\footnote{Even when the document structure is defined using XML 
Schema, documents may contain elements prefixed with other namespaces.}. The type of \ident{elem}
is thus a \emph{variant} type annotated with the three statically known possibilities (heading,
paragraph, image). The variant is mapped to the following F\# type:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{Element}~=  \\[0.1em]
 \quad \kvd{member}~\ident{Heading}~:~\ident{option}\langl \ident{string} \rangl\\
 \quad \kvd{member}~\ident{Paragraph}~:~\ident{option}\langl \ident{string} \rangl\\
 \quad \kvd{member}~\ident{Image}~:~\ident{option}\langl \ident{Image} \rangl\\
\end{array}
\end{equation*}
%
The type provides a nice access to the elements known statically from the sample, but it encodes
the \emph{open-world assumption} -- the user also needs to handle the case when the value is none 
of the statically known elements.

One perhaps surprising consequence of the use of variant types is that the above code will work
correctly on document that contains elements other than those in the sample document. This would 
arguably not be desirable design for data types in a programming language, but it is extremely
useful when reading data from external data sources. It is also worth noting that our type 
inference attempts to minimze the number of variant types that appear in the resulting type --
a variant is not used if there is another option.

@@

\subsection{Summary}
The previous three sections demonstrated the F\# Data type providers for JSON, CSV and XML.
A reader interested in more examples is invited to look at documentation on the F\# Data web 
site\footnote{Available at \url{http://fsharp.github.io/FSharp.Data/}}.

It should be noted that we did not attempt to present the library using well-structured ideal 
samples, but instead used data sets that demonstrate typical issues that are frequent in real 
world inputs (missing data, inconsistent encoding of Booleans and heterogeneous structures).

From looking at just the previous three examples, these may appear arbitrary, but our experience
suggests that they are the most common issues. The following JSON response with government debt
information returned by the World Bank API demonstrates all three problems:
%
{\small{
\begin{verbatim}
  [ { "page": 1, "pages": 5 },
    [ { "indicator": "GC.DOD.TOTL.GD.ZS",
        "date": "2012",
        "value": null },
      { "indicator": "GC.DOD.TOTL.GD.ZS",
        "date": "2010",
        "value": "35.1422970266502" } ] ]
\end{verbatim}
}}
%
\noindent
First of all, the top-level element is a collection containing two values of different kind.
The first is a record with meta-data about the current page and the second is an array with data. 
The JSON type provider infers this as heterogeneous collection with properties \ident{Record}
for the former and \ident{Array} for the latter. Second, the \ident{value} is \kvd{null} for some
records. Third, numbers can be represented in JSON as numeric literals (without quotes), but
here, they are returned as string literals instead\footnote{This is often used to avoid non-standard
numerical types in JavaScript.}.

@@

\newpage
~
\newpage
\section{Structural type inference}
\label{sec:inference}

Our type inference algorithm for structured formats is based on a subtyping relation. When 
inferring the type of a document, we infer the most specific types of individual values (CSV rows,
JSON or XML nodes) and recursively find a common supertype of all child nodes or sample documents.

In this section, we define the \emph{structural type} $\sigma$, which is a type of structured data.
Note that this type is distinct from the programming language types $\tau$ (type providers generate 
the latter from the former). Next, we define the subtyping relation on structural types $\sigma$ and 
describe the algorithm for finding a common supertype.

@@

We distinguish between \emph{non-nullable types} that always have a valid value (written as 
$\hat{\sigma}$) and \emph{nullable types} that encompass missing and \kvd{null} values 
(written as $\sigma$). We write $\nu$ for record field names and $\nu_?$ for names that can be empty:
%
\begin{equation*}
\begin{array}{rcl}
 \hat{\sigma} &\narrow{=}& \nu_?\; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \} \\[0.1em]
                &\narrow{|}& \ident{float} \lsep \ident{decimal} \lsep \ident{int} \lsep \ident{bool} \lsep \ident{string} 
 \\[0.6em] 
       \sigma &\narrow{=}& ~\hat{\sigma}~ \lsep \kvd{option}\langl \hat{\sigma} \rangl \lsep [\sigma] \\[0.1em]
              &\narrow{|}& \kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl \lsep ~\bot~ \lsep \kvd{null}
\end{array}
\end{equation*}
%
Non-nullable types include records (consisting of an optional name and fields with their types) 
and primitive types. Records arising from XML documents are named, while records used by 
JSON and CSV providers are unnamed.

@@

Any non-nullable type can be wrapped in the \kvd{option} constructor to explicitly permit the 
\kvd{null} value. Type providers map \kvd{option} to standard F\# option type. A collection 
type $[\sigma]$ is also nullable and missing values or \kvd{null} are treated as empty collection. 
The type $\kvd{null}$ is inhabited by the $\kvd{null}$ value (using an overloaded notation) and 
$\bot$ is the bottom type.

>>>>

\documentclass[10pt,preprint,blind,clearpagebib]{sigplanconf}

\usepackage{listings}
\usepackage{enumitem}
\usepackage[hyphens]{url}
\usepackage[svgnames]{xcolor}
\definecolor{lcl}{RGB}{140,0,100}
\usepackage[colorlinks=true,breaklinks,draft=false]{hyperref}
\hypersetup{urlcolor=lcl,linkcolor=lcl,citecolor=lcl}
\newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}
\usepackage{amssymb}
\usepackage{multicol}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage[fleqn]{amsmath}
\usepackage{graphics} 
\usepackage{stmaryrd}
\usepackage{amsthm}
\usepackage{ftnright}
\usepackage[T1]{fontenc}
\usepackage{semantic}
\usepackage{enumitem}
\usepackage[hang,flushmargin]{footmisc} 
\usepackage[notquote]{hanging}
\usepackage{flushend}

@@

% TODO: Discuss 'stability' of the inference (minor change in schema does not cause dramatic cahnges)
% TODO: Say that variant is a last-resort thing
% TODO: Practical evaluation (lots of downloads, lots of Github - how often people customize samples?)
% TODO: Explain why we need option type (because T + null is a variant)
% TODO: In the PLDI version, we delete all the unnecessary runtime conversions
% but we should say that those happen for practical reasons (e.g. float -> int, or anyhing to string)

@@

\begin{abstract}
Most modern applications interact with external services and access data in structured formats such 
as XML, JSON and CSV. Static type systems do not understand such formats making accessing data more 
cumbersome. Should we give up and leave the messy world of external data to dynamic typing and 
runtime checks? Of course, not!

@@

For example, OpenWeatherMap provides an end-point for getting current weather for a 
given city\footnote{See ``Current weather data'': \url{http://openweathermap.org/current}}. 
The documentation describes the URL parameters and shows one sample JSON to illustrate the typical
response structure. Using a standard library for working with JSON and HTTP, we might call the 
service and read the temperature as follows:
%
\begin{equation*}
\begin{array}{l}
 \kvd{let}~\ident{doc}=\ident{Http.Request}(\str{http://api.owm.org/?q=NYC}) \\
 \kvd{match}~\ident{JsonValue.Parse}(\ident{doc})~\kvd{with} \\
 |~\ident{Record}(\ident{root})\rightarrow \\
 \quad \kvd{match}~\ident{Map.find}~\str{main}~\ident{root}~\kvd{with} \\
 \quad |~\ident{Record}(\ident{main})\rightarrow \\
 \quad \quad \kvd{match}~\ident{Map.find}~\str{temp}~\ident{main}~\kvd{with} \\
 \quad \quad |~\ident{Number}(\ident{num})\rightarrow \ident{printfn}~\str{Nice \%f degrees!}~\ident{num} \\
 \quad \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\
 \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\
 |~\_\rightarrow \ident{failwith}~\str{Incorrect format} 
\end{array}
\end{equation*}
%
The code assumes that the response has a particular format described in the documentation. The
root node must be a record with a \str{main} field, which has to be another record containing
a numerical \str{temp} field. When the format is incorrect, the data access simply fails
with an exception.

Using the JSON type provider from F\# Data, we can write code with exactly the 
same functionality in two lines:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{W} = \ident{JsonProvider}\langl\str{http://api.owm.org/?q=NYC}\rangl \\[0.1em]
 \ident{printfn}~\str{Nice \%f degrees!}~(\ident{W.GetSample().Main.Temp})
\end{array}
\end{equation*}
%
$\ident{JsonProvider}\langl\str{...}\rangl$ invokes a type provider at 
compile-time with the URL as a sample. The type provider infers the structure of the response
and provides a type with a \ident{GetSample} method that returns a parsed JSON with nested
properties \ident{Main.Temp}, returning the temperature as a number.

@@

@@

We start with an informal overview that shows how F\# Data type providers simplify working with 
JSON, XML and CSV. We introduce the necessary aspects of F\# type providers along the 
way. The examples in this section illustrate a number of key properties of our type inference 
algorithm:

\begin{itemize}
\item Our mechanism is predictable. The user directly works with the inferred types and should 
  understand why a specific type was inferred from a given sample\footnote{In particular, we do 
  not use probabilistic methods where adding additional sample could change the shape of the type.}.

\item Our inference mechanism prefers records. This supports developer tooling --
  most F\# editors provide code completion hints on ``.'' and so types with properties (records)
  are easier to use than types that require pattern matching.

\item Finally, we handle a number of practical concerns important in the real world. This includes 
  support for different numerical types, \kvd{null} values and missing data.
\end{itemize}

@@

The JSON format used in \S\ref{sec:introduction} is a populardata exchange format based on 
JavaScript data structures. The following is the definition of \ident{JsonValue} 
used earlier to represent JSON:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{JsonValue} = \\[0.1em]
 \quad|~ \ident{Number}~\kvd{of}~\ident{decimal} ~~|~ \ident{Null} \\[0.1em]
 \quad|~ \ident{String}~\kvd{of}~\ident{string} \qquad\, |~ \ident{Boolean}~\kvd{of}~\ident{bool} \\[0.1em]
 \quad|~ \ident{Record}~\kvd{of}~\ident{Map}\langl\ident{string}, \ident{JsonValue}\rangl \\[0.1em]
 \quad|~ \ident{Array}~\kvd{of}~\ident{JsonValue}[] \\[0.1em]
\end{array}
\end{equation*}
%
The earlier example used only a nested record containing a number. To demonstrate other 
aspects of the JSON type provider, we look at an example that also involves an array:
%
{\small{
\begin{verbatim}
  [ { "name":"Jan", "age":25 }, { "name":"Tomas" },
    { "name":"Alexander", "age":3.5 } ]
\end{verbatim}
}}
%
\noindent
Say we want to print the names with an age if it is available. The standard approach would 
be to pattern match on the parsed \ident{JsonValue}, check that the top-level node is a \ident{Array}
and iterate over the elements checking that each is a \ident{Record} with certain properties. We
would throw an exception or skip values in incorrect format. Even if we defined helper functions, 
we would still need to specify field names as strings, which is error prone and can not be 
statically checked.

Assuming \strf{people.json} is the above example and \ident{data} is a string containing
JSON in the same format, we can write:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{People}~=~\ident{JsonProvider}\langl\str{people.json}\rangl\hspace{1em} \\[0.6em]
 \kvd{for}~\ident{item}~\kvd{in}~\ident{People.Parse}(\ident{data})~\kvd{do}\\[0.1em]
 \quad\ident{printf}~\str{\%s }~\ident{item.Name}\\[0.1em]
 \quad\ident{Option.iter}~(\ident{printf}~\str{(\%f)})~\ident{item.Age}
\end{array}
\end{equation*}
%
The code achieves the same simplicity as when using dynamically typed languages, but it is statically 
type-checked. In contrast to the earlier example, we now use a local file \strf{people.json} as a 
sample for the type inference, but then processes data from another source.

\paragraph{Type providers.}
The notation $\ident{JsonProvider}\langl\str{people.json}\rangl$ passes a \emph{static parameter} 
to the type provider. Static parameters are resolved at compile-time and have to be constant. 
The provider analyzes the sample and generates a type that we name \ident{People}. In most F\# 
editors, the type provider is also executed at development-time and the provided types are 
used in auto-completion.

The \ident{JsonProvider} uses a type inference algorithm (\S\ref{sec:inference})  and
infers the following types from the sample:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{Entity}~=  \\
 \quad \kvd{member}~\ident{Name}~:~\ident{string} \\
 \quad \kvd{member}~\ident{Age}~:~\ident{option}\langl \ident{decimal}\rangl \\[0.5em]
 \kvd{type}~\ident{People}~=  \\
 \quad \kvd{member}~\ident{GetSample}~:~\ident{unit}~\rightarrow~\ident{Entity}[] \\
 \quad \kvd{member}~\ident{Parse}~:~\ident{string}~\rightarrow~\ident{Entity}[] \\
\end{array}
\end{equation*}
%
The type \ident{Entity} represents the person. The field \ident{Name} is available for all
sample values and is inferred as \ident{string}. The field \ident{Age} is marked as optional,
because the value is missing in one sample. The two sample ages are an integer $25$ and a 
decimal $3.5$ and so the common inferred type is \ident{decimal}.

The type \ident{People} has two methods for reading data. \ident{GetSample} parses the
sample used for the inference and \ident{Parse} parses a JSON string containing data in the same 
format as the sample. Both methods return an array of \ident{Entity} values.

\paragraph{Error handling.}
In addition to the structure of the types, the type provider also specifies what code should be 
executed at run-time in place of \ident{item.Name} and other operations. The runtime behaviour is 
the same as in the earlier hand-written sample (\S\ref{sec:introduction}). A member access 
throws an exception if the document does not have the expected format.

Informally, the safety property (\S\ref{sec:safety}) states that if the inputs
are subtypes of some of the provided samples (i.e.~the samples are representative), then
no exceptions will occur. In other words, we cannot avoid all failures, but we can prevent some.
If the OpenWeatherMap changes the response format, the sample in \S\ref{sec:introduction}
will not re-compile and the user knows that the code needs to be changed. What this means for
the traditional ML type safety is discussed in \S\ref{sec:safety-discuss}.

\paragraph{The role of records.}
The sample code is easy to write thanks to the fact that most F\# editors provide code completion
when ``.'' is typed. The developer does not need to look at the sample JSON file to see what fields
are available for a person. To support this scenario, our inference algorithm prefers records 
(we also treat XML elements and CSV rows as records) and only infers types that require pattern 
matching as the last resort.

@@

\subsection{Processing XML documents}
\label{sec:providers-xml}

XML documents are formed by nested elements with attributes. We can view elements as records with 
a field for each attribute and an additional special field for the nested contents (which is a 
collection of elements).

Consider a simple extensible document format where a root element
{\ttfamily\small <doc>} can contain a number of document element, one of which is
{\ttfamily\small <heading>} representing headings:
%
{\small{
\begin{verbatim}
  <doc>
    <heading>Working with JSON</heading>
    <p>Type providers make this easy.</p>
    <heading>Working with XML</heading>
    <p>Processing XML is as easy as JSON.</p>
    <image source="xml.png" />
  </doc>
\end{verbatim}
}}
%
\noindent
The F\# Data library has been designed primarilly to simplify reading of data. For example,
say we want to print all headings in the document. The sample shows a part of the document structure 
(in particular the {\ttfamily\small <heading>} element), but it does not show all possible elements 
(say, {\ttfamily\small <table>}). Assuming the above document is \strf{sample.doc}, we can write:
%
\noindent
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{Document}~=~\ident{XmlProvider}\langl\str{sample.doc}\rangl\hspace{1em} \\[0.5em]
 \kvd{let}~\ident{root}~=~\ident{Document.Load}(\str{http://.../another.doc})\\
 \kvd{for}~\ident{elem}~\kvd{in}~\ident{root.Doc}~\kvd{do}\\
 \quad\kvd{match}~\ident{elem.Heading}~\kvd{with} \\
 \quad|~\ident{Some text} \rightarrow \ident{printf}~\str{ - \%s }~\ident{text}\\
 \quad|~\ident{None} \rightarrow ()\\
\end{array}
\end{equation*}
%
The example iterates over a collection of elements \ident{elem} returned by \ident{root.Doc}. 
The type provides a typed access to elements known from the sample document and so we can
write \ident{elem.Heading}, which returns an optional value.

\paragraph{Open world and variant types.}
In the above example, we do not infer a union type representing a choice between 
heading, paragraph and image. By its nature, XML is extensible and the sample cannot include all 
possible nodes\footnote{Even when the document structure is defined using XML Schema, documents may 
contain elements prefixed with other namespaces.}. The type of \ident{elem} is thus a top type (we 
call it \emph{variant}) annotated with the statically known possibilities (heading, paragraph, image). 
It is mapped to the following F\# type:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{Element}~=  \\[0.1em]
 \quad \kvd{member}~\ident{Heading}~:~\ident{option}\langl \ident{string} \rangl\\
 \quad \kvd{member}~\ident{Paragraph}~:~\ident{option}\langl \ident{string} \rangl\\
 \quad \kvd{member}~\ident{Image}~:~\ident{option}\langl \ident{Image} \rangl\\
\end{array}
\end{equation*}
%
The type provides a nice access to the elements known statically from the sample, but it encodes
the \emph{open-world assumption} -- the user also needs to handle the case when the value is none 
of the statically known elements.

As a consequence, the variant type can be seen as the top type. The above code will work
correctly on document that contains any elements, including those not listed in the sample. This
is extremely useful when reading data from external data sources, because the sample document only
needs to contain values relevant to the application. It is also worth noting that our type 
inference attempts to minimze the number of variant types that appear in the resulting type --
a variant is not used if there is another option (Corollary~\ref{thm:no-unions}).

@@

\subsection{Summary}
In addition to JSON and XML (\S\ref{sec:providers-json}, \S\ref{sec:providers-xml}), F\# Data also 
implements a type provider for CSV. We treat CSV data as lists of records (with field for each 
column) and so handling of CSV is just a special case of our inference algorithm.

Throughout the introduction, we used data sets that demonstrate typical issues that are frequent
in real-world inputs (missing data, inconsistent encoding of primitive values and heterogeneous 
structures). In our experience, these are the most common issues. The following JSON response with 
government debt information returned by the World Bank API\footnote{Available at 
\url{http://data.worldbank.org}} demonstrates all three problems:
%
{\small{
\begin{verbatim}
  [ { "page": 1, "pages": 5 },
    [ { "indicator": "GC.DOD.TOTL.GD.ZS",
        "date": "2012", "value": null },
      { "indicator": "GC.DOD.TOTL.GD.ZS",
        "date": "2010", "value": "35.14229" } ] ]
\end{verbatim}
}}
%
\noindent
First of all, the top-level element is a collection containing two values of different kind.
The first is a record with meta-data about the current page and the second is an array with data. 
The actual implementation supports a concept of heterogeneous collections (omitted in this paper
for space reasons) and provides a type with properties \ident{Record} for the former and 
\ident{Array} for the latter. Second, the \ident{value} is \kvd{null} for some records. Third, 
numbers can be represented in JSON as numeric literals (without quotes), but here, they are 
returned as string literals instead\footnote{This is often used to avoid non-standard numerical 
types of JavaScript.}.

@@

\section{Structural type inference}
\label{sec:inference}

Our type inference algorithm for structured formats is based on a subtyping relation. When 
inferring the type, we infer the most specific types of individual values (CSV rows, JSON or XML 
nodes) and recursively find a common supertype of all child nodes or all sample documents.

We first define the type of structured data $\sigma$. Note that this type is distinct 
from the programming language types $\tau$ (type providers generate the latter from the former). 
Next, we define the subtyping relation on structural types $\sigma$ and describe the algorithm 
for finding a common supertype.

@@

We distinguish between \emph{non-nullable types} that always have a valid value (written as 
$\hat{\sigma}$) and \emph{nullable types} that encompass missing and \kvd{null} values 
(written as $\sigma$). We write $\nu$ for record field names and $\nu_?$ for names that can be empty:

\noindent
\begin{equation*}
\begin{array}{rcl}
 \hat{\sigma} &\narrow{=}& \nu_?\; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \} \\[0.1em]
                &\narrow{|}& \ident{float} \lsep \ident{decimal} \lsep \ident{int} \lsep \ident{bool} \lsep \ident{string} 
 \\[0.6em] 
       \sigma &\narrow{=}& ~\hat{\sigma}~ \lsep \kvd{option}\langl \hat{\sigma} \rangl \lsep [\sigma] \\[0.1em]
              &\narrow{|}& \kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl \lsep ~\bot~ \lsep \kvd{null}
\end{array}
\end{equation*}
%
Non-nullable types include records (consisting of an optional name and fields with their types) 
and primitive types. Records arising from XML documents are named, while records used by 
JSON and CSV providers are unnamed.

@@

Any non-nullable type can be wrapped in the \kvd{option} constructor to explicitly permit the 
\kvd{null} value. Type providers map \kvd{option} to standard F\# option. A collection $[\sigma]$ 
is also nullable and missing values or \kvd{null} are treated as empty collections. The type 
$\kvd{null}$ is inhabited by the $\kvd{null}$ value (using an overloaded notation) and $\bot$ 
is the bottom type.

A variant type \kvd{any} is annotated with the types it may represent in the sample document.
As discussed earlier (\S\ref{sec:providers-xml}), variants force the user to handle the case when 
a value is of a different type than the statically known ones and so it can be seen as a family of 
top types. For the same reason, variants also implicitly permit the \kvd{null} value.
